Supplementary Information

### 2021-02-08 doc/manuscript/MEE_manuscript.Rmd
minor fix

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Feb 8)
n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
n_spst_RW_DBH_sig=35 #  species with sig first-order linear DBH term
n_spst_RW_DBH_2nd_order=26
n_spst_RW_DBH_2nd_order_neg=21
n_spst_BAI_DBH_best_model=42 #exception: ACRU at HF  
n_spst_BAI_DBH_best_model_pos=40 #exceptions: PIPO at LT (reconstructed DBH down to only 15cm) and PILO at CB (short DBH range on old trees), both neg but not significant
n_spst_BAI_DBH_best_model_2nd=41
n_spst_BAI_DBH_best_model_2nd_neg=40 # exception: PILO at CB
n_spst_dAGB_DBH_best_model=42 # exception: PILO at CB
```

**All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. `r size`).** 
$RW$ varied significantly with $DBH$ for `r round(n_spst_RW_DBH_sig/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_sig` of `r n_spst_DBH_analyzed`), and the majority of best models included a significant second-order linear $DBH$ term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative). 
There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
On one end of the spectrum, some species exhibited maximum $RW$ at low $DBH$, followed by fairly rapid declines in $RW$ with increasing $DBH$. 
Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) and/or grew in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig. `r size`).
At the other end of the spectrum, some species had low $RW$ at small $DBH$, increased to peak $RW$ at intermediate $DBH$, and subsequently declined.
These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp. at SCBI and Žofín, *Acer saccharum* at LDW, *Picea* spp. at Žofín and CB; Table `r species_table`).

@@

Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ for the vast majority of site-species combinations (`r round(n_spst_RW_DBH_sig/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig. `r size`). 
Variation in these patterns--particularly for $RW$, for which variation was most pronounced--was driven by two primary, interrelated factors: species ecology and stand history.
Species that would have established in fairly open conditions--*i.e.*, shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth, measured as $RW$, followed by a roughly exponential decline.
Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments [@fritts_tree_1976; @biondi_theory-driven_2008].
However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, $RW$.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (*i.e.*, a "cross-sectional" analysis), $RW$ increases continuously with $DBH$ [e.g., @muller-landau_testing_2006; @anderson-teixeira_size-related_2015; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_species-specific_2018].

>>>>

```{r , eval=TRUE, echo=FALSE}
## Tallies for the  paragraph below (Feb 8)
n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
n_spst_RW_DBH_bm=35 #  species with first-order linear DBH term in best model
n_spst_RW_DBH_2nd_order=26
n_spst_RW_DBH_2nd_order_neg=21
n_spst_BAI_DBH_best_model=42 #exception: ACRU at HF  
n_spst_BAI_DBH_best_model_pos=40 #exceptions: PIPO at LT (reconstructed DBH down to only 15cm) and PILO at CB (short DBH range on old trees), both neg but not significant
n_spst_BAI_DBH_best_model_2nd=41
n_spst_BAI_DBH_best_model_2nd_neg=40 # exception: PILO at CB
n_spst_dAGB_DBH_best_model=42 # exception: PILO at CB
```

**All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. `r size`).** 
For $RW$, $DBH$ was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear $DBH$ term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative). 
There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
On one end of the spectrum, some species exhibited maximum $RW$ at low $DBH$, followed by fairly rapid declines in $RW$ with increasing $DBH$. 
Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) and/or grew in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig. `r size`).
At the other end of the spectrum, some species had low $RW$ at small $DBH$, increased to peak $RW$ at intermediate $DBH$, and subsequently declined.
These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp. at SCBI and Žofín, *Acer saccharum* at LDW, *Picea* spp. at Žofín and CB; Table `r species_table`).

@@

Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ for the vast majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig. `r size`). 
Variation in these patterns--particularly for $RW$, for which variation was most pronounced--was driven by two primary, interrelated factors: species ecology and stand history.
Species that would have established in fairly open conditions--*i.e.*, shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth, measured as $RW$, followed by a roughly exponential decline.
Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments [@fritts_tree_1976; @biondi_theory-driven_2008].
However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, $RW$.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (*i.e.*, a "cross-sectional" analysis), $RW$ increases continuously with $DBH$ [e.g., @muller-landau_testing_2006; @anderson-teixeira_size-related_2015; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_species-specific_2018].
